---
title: "Reproducible Research: Peer Assessment 1"
author: Yukai Zou
date: "11 November 2015"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

We began with setting the working directory and loading the data:
```{r section 1}
setwd("~/Desktop/Data Sciences Specialization/Reproducible Research/RepData_PeerAssessment1/")
activity <- read.csv("activity.csv", stringsAsFactors=FALSE)
head(activity)
```


## What is mean total number of steps taken per day?

To answer the question, first we calculated the total number of steps taken per day:
```{r section 2.1}
ans1 <- tapply(activity$steps, activity$date, sum)
ans1
```
We then made a histogram to observe the distribution:
```{r section 2.2}
par(mar = c(4, 4, 2, 2))
hist(ans1, main = "Distribution of the total number of steps taken each day", 
     xlab = "Number of steps", col = "lightgreen")
```

The distribution is somewhat skewed to the left, but seems that the majority lies between 10000 and 15000. We then calculated the mean and median of the total number of steps taken per day:
```{r section 2.3}
c(mean = mean(ans1, na.rm=TRUE), median = median(ans1, na.rm=TRUE))
```
The median is slightly lower than the mean value.


## What is the average daily activity pattern?

To answer this question, we first averaged the number of steps, ignoring all the missing values, within each interval; then we plotted a time-series line graph with the 5-minute interval as x axis and the averaged number as y-axis.
```{r section 3.1}
ans2 <- tapply(activity$steps, activity$interval, function(x) mean(x, na.rm=TRUE))
plot(ans2, type="l", xlab = "Index of interval", ylab = "Averaged number of steps",
     main = "Averaged steps in each 5-minute interval")
```

Based on the time-series plot, a peak was identified around the 100th interval. We then applied `which()` to find which index the peak value corresponds to:
```{r section 3.2}
c(interval = names(which(ans2 == max(ans2))), index = as.numeric(which(ans2 == max(ans2))))
```
Therefore, we found that on average across all the days, the maximum number of steps was recorded at the **104th** interval, from **835** to **840 minutes**.

## Imputing missing values

We suspected that the presence of missing values (e.g. missing days/intervals, indicated by `NA`) may introduce bias in our calculations and summaries, therefore, it is necessary to impute these missing values. First, we wanted to know how many missing values in total are there within the dataset:
```{r section 4.1}
missing <- is.na(activity$steps)
sum(missing)
```
That's quite a few. To impute the missing data, the mean of each 5-minute interval was used to fill in all of the missing values in the dataset, and a new dataset was created which contains the same original dataset but with the missing data filled in:
```{r section 4.2, cache=TRUE}
activity2 <- activity
interval <- activity2[missing,]$interval
for (i in 1:length(interval)) {
        activity2[missing,]$steps[i] <- ans2[which(names(ans2) == interval[i])] 
        # impute the missing date with the corresbonding mean value
}
head(activity2)
```
The imputation has done successfully. We then made a histogram to look at the new distribution of the total number of steps taken each day:
```{r section 4.3}
ans3 <- tapply(activity2$steps, activity2$date, sum)
hist(ans3, main = "Distribution of the total number of steps taken each day", xlab = "Number of steps", 
     col = "lightblue")
```

The distribution, after imputing the missing data, looks less skewed and more concentrated around the mean. We then calculated the new mean and median of the total number of steps taken per day:
```{r section 4.4}
c(mean = mean(ans3), median = median(ans3))
```
Compared with the previous numbers, imputing missing data did not alter the mean, and the median became identical with the mean.

## Are there differences in activity patterns between weekdays and weekends?

To compare the activity patterns between weekdays and weekends, a new factor variable was added to the dataset using *dplyr* package, indicating whether a given date is a weekday or weekend day.
```{r section 5.1}
library(dplyr)
activity2$date <- as.Date(activity2$date)
activity2 <- mutate(activity2, daytype = factor(1 * (weekdays(date) == "Saturday" | 
                     weekdays(date) == "Sunday"), labels = c("weekday", "weekend")))
```
We then made a line graph containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken (y-axis), averaged across all weekday days (red line) or weekend days (blue line). The two lines were overlayed to make the comparison more intuitive.
```{r section 5.2}
weekday <- split(activity2, activity2$daytype)[[1]] # Subset the weekday days
weekend <- split(activity2, activity2$daytype)[[2]] # Subset the weekend days
weekday.avg <- tapply(weekday$steps, weekday$interval, mean) # Average across weekday days
weekend.avg <- tapply(weekend$steps, weekend$interval, mean) # Average across weekend days
plot(weekday.avg, type="l", col = "red", main = "Average number of steps by weekdays and weekends", 
     xlim = c(0,300), xlab = "Index of 5-minute interval", ylab = "Average number of steps")
lines(weekend.avg, col = "blue")
legend("topright", legend = c("weekdays", "weekends"), col = c("red", "blue"), lty = 1, cex = 1)
```

From the graph, we can see that there is not much difference in the first 50 intervals and after 250th interval between the two lines. At the first half of the day (from the 50th to around 130th interval), the average number of steps at weekdays is overall higher than that at weekends, whereas at the second half of the day (from the 130th to the 250th interval), the average number of steps was recorded with a relatively higher value at weekends.